Supernovae Time Profiles as a Portal to New Physics at Neutrino Telescopes

### 2024-01-18 paper/sections/conclusion.tex
Updates from Overleaf

\textbf{\textit{Results}}---
\begin{figure}[t!]
    \centering
    \includegraphics[width=0.47\textwidth]{figures/majoran_sensitivity}
    \caption{\textbf{\textit{Exclusion limit for the majoron case.}}
    }
    \label{fig:sensitivity}
\end{figure}
We present the exclusion limit in \fig{sensitivity}.

\textbf{\textit{Outlook}}---

>>>>

\textbf{\textit{Results}}---
\begin{figure}[t!]
    \centering
    \includegraphics[width=0.47\textwidth]{figures/majoran_sensitivity}
    \caption{\textbf{\textit{Exclusion sensitivities for the Majoron case.}}
    The lines on this plot show the exclusion sensitivity for IceCube, and two next-generation neutrino experiments, Hyper-Kamiokande and DUNE.
    Additionally, we also show the reaches by SN cooling (blue) and by the lack of observation of high-energy neutrinos from SN1987A (grey) \cite{Fiorillo:2022cdq}, with the width of the lines quantifying uncertainties from SN modeling.
    }
    \label{fig:sensitivity}
\end{figure}
Looking for rapid raises within various time windows at neutrino telescope in the occurrence of future nearby SN explosion will provide exceptional tests of the parameter space. Assuming the SN event happens in the galaxy at a distance $D_{\rm SN}=\unit[10]{kpc}$, which is not unlikely~\cite{Reed:2005en,Rozwadowska:2020nab}, we consider the currently operating IceCube neutrino observatory as an example. 
The expected exclusion limit are shown in \cref{fig:sensitivity} (dark red) if no excess of hits on top of those from background noise and the standard neutrino flux were observed. JUNO expected to start taking data soon, while DUNE and Hyper-K requiring longer time to start operating, will be able to test the parameter of Majorons by looking for high energy neutrino following \cite{Fiorillo:2022cdq} if such SN event happens during their operation time. We present the estimated reaches of JUNO, DUNE and Hyper-K in \cref{fig:sensitivity} as xxx, red and orange lines, respectively, leaving aside careful detector analysis, together with current limits \cite{Fiorillo:2022cdq} from the aforementioned energy loss requirement (blue band), and non-observation of high energy neutrinos from SN 1987A (gray band) in \cref{fig:sensitivity}.

We observe that while for the intermediate mass region, the constraints from IceCube are comparable to that from DUNE and Hyper-K, at both low mass region $m_\phi \lesssim \unit[10]{MeV}$ and high mass region $m_\phi \gtrsim \unit[200]{MeV}$, IceCube can provide stronger constraints. This is the outcome we expect. For light $\phi$ case with negligible time delay, the neutrino signals can  arrive at the detector even before the peak of standard neutrino flux arrives. The resulting hits would appear in a time window separated from that when most of standard flux contribution appears, as we show in the left panel of \cref{fig:hits_and_likelihood}, thus enhancing the reaches. On the high mass end, neutrino signals will arrive much later than the standard case, where a time window at late time can be considered to reduce the standard neutrino contaminations.

Uncertainties on the analysis from different modeling SN explosion are minor, as studies in \cite{li2023old} and potential discrepancies between data and modeling are unresolved questions. Nevertheless we point out that the discrepancy could be within $2\sigma$ level which will only change our exclusion limit by a small factor.

\textbf{\textit{Outlook}}---
we can also study axion cases, mixing cases with this strategy. Pre-supernova bouncing.

\begin{acknowledgments}
We would like to thank xxx for stimulating discussions about this work. Y.-Y. L is supported by the NSF of China through Grant
No. 12247103 and Grant No. xxx.
\end{acknowledgments}

### 2024-01-18 paper/sections/detector_response.tex
Updates from Overleaf

\textbf{\textit{Detector Response and Statistical Treatment}}---In order to simulate the

\begin{figure*}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/hits_and_likelihood.pdf}
    \caption{\textbf{\textit{Test statistic as a function of $t_{\mathrm{start}}$ and $\Delta$t.}
    }}
    \label{fig:hits_and_likelihood}
\end{figure*}

>>>>

\textbf{\textit{Detector Response and Statistical Treatment}}---The IceCube Neutrino Observatory comprises 5,160 light-detecting digital optical modules (DOMs) buried in a cubic kilometer of the deep, transparent Antarctic ice sheet.
When neutrinos interact within or near the detector, the charged by-products emit photons that the DOMs can detect.
The DOMs are arranged on 86 strings of 60 DOMs with an inter-string distance between 70~m and 125~m.
This enables IceCube to resolve individual neutrinos with energies approximately ranging between a few GeV and a few PeV.
The neutrinos produced by SNe are far below this threshold, and the neutrinos cannot be individually resolved; however, the immense number of neutrinos produced in a SN produces an increase in the single-photon rate of the detector.
This increase in the rate of photons can be distinguished by the background caused by dark noise and radioactive activity in the DOM glass to enable the detection of galactic SNe events~\cite{Griswold:2023iwz}.

New neutrino physics will affect the development of a SN, and will distort the temporal structure of the photon signal seen in the IceCube detector.
Thus, one may look for excess events in certain time intervals as evidence of this new physics.
In this work, we simulate the light curve produced by a standard SN with a progenitor mass of $8.8~M_{\odot}$ and those produced by different BSM scenarios; see the left-hand panel of Fig.~\ref{fig:hits_and_likelihood} for example light curves.
We use the \texttt{ASTERIA}~\cite{spencer_griswold_2020_3926835} package in order to simulate the detector response to the different SNe scenarios as well as the thermal and radioactive noise in the DOMs. \yy{we need to provide a brief recasting of simulating the detector response, especially for different flavors.}
This package simulates light yields from coherent $\nu_{e}$ scattering off electrons in the ice, inverse beta decay of $\bar{\nu}_{e}$ on nuclei in the ice, and charged- and neutral-current interactions for $\nu_{\alpha}$.
In order to interface to this package, we use \texttt{SNEWPY}~\cite{baxter2021snewpy} package and, in particular, the \texttt{ParametrizedFlux} object.

With the number of hits in the detector from noise, SM-only scenarios, and BSM scenarios as a function of time, we can then quantify the probability of seeing a certain number of hits in a given time window with:
\begin{equation}
    -2\Delta\mathrm{LLH} = 2 \left[N_{\mathrm{exp.}} - N_{\mathrm{obs.}} + N_{\mathrm{obs.}}\log\left(\frac{N_{\mathrm{obs.}}}{N_{\mathrm{exp.}}}\right)\right].
    % -2\Delta\mathrm{LLH}(t_{\mathrm{start}}, \Delta t) = 2 \left[N_{\mathrm{exp.}} - N_{\mathrm{obs.}} + N_{\mathrm{obs.}}\log\left(\frac{N_{\mathrm{obs.}}}{N_{\mathrm{exp.}}}\right)\right].
\end{equation}
where $N_{\mathrm{obs.}}$ is the number of photons seen in the detector in the given time given and $N_{\mathrm{exp.}}$ is the number of photons expected from a particular BSM hypothesis.
In the right-hand panel of Fig.~\ref{fig:hits_and_likelihood}, we show the likelihood space as a function of the time at which the time range begins, $t_{\mathrm{start}}$ and the duration of the time range, $\Delta t$ for an example BSM hypothesis.

For each physics hypothesis, we then select the time range that maximizes the test statistic and use that maximal test statistic value.
The optimal ranges for each physics hypothesis are shown by the dashed lines and arrows in the left panel of Fig.~\ref{fig:hits_and_likelihood}.

### 2024-01-18 paper/sections/model.tex
Updates from Overleaf

\textbf{\textit{The Models}}---
As an example, we will consider the Majoron model. Nevertheless the study can also be applied to sterile neutrinos via magnetic portal. For the case of sterile neutrino with mixing portal, special care needs to be taken for resonant conversions in which case active neutrinos with plasma mass would travel as slowly as the sterile neutrinos at specific distances.
The relevant part of the Lagrangian in the Majoron case reads
\begin{align}
\mathcal{L} \supset -\frac{g_{\alpha\beta}}{2} \nu_\alpha \nu_\beta \phi - M_\phi \phi\phi^* + \text{h.c.}\,,
\end{align}
where $M_\phi$ is the Majoron mass and $g$ parametrizes interaction strength between $\phi$ and neutrinos. Consider flavor universal interaction, we abbreviate $g_{\alpha\beta}\equiv g$. Majorons are produced from neutrino coalescence in the star and then subsequently decay to a pair of neutrinos with the total decay width to all flavor of neutrinos given by $\Gamma_\phi = 3g^2 m_\phi/16 \pi$, and as a result, BSM neutrino flux is produced.

Using data simulated by the Garching group for an $8.8 M_\odot$ progenitor star \cite{Huedepohl2010}, we obtained the fluxes of Majoron $\phi$, and produced the cooling limit, which is in agreement with those obtained in literature \cite{Fiorillo:2022cdq}.
As mentioned above, for the regions below the coupling strength, Majorons produced will immediately stream out. Along its way towards the Earth, Majorons will decay to active neutrinos. For a given emission angle $\alpha$ of daughter neutrino relative to the direction of mother particle, and the angle $\theta$ at which the daughter neutrino arrives at the Earth, the time delay $\Delta t$, the arrival time of the daughter $\nu$ relative to that of the neutrinos produced in the explosion via SM processes, is \cite{Jaeckel:2017tud}
\begin{align}
    \Delta t = L_1/\beta + D_{\rm SN} \cos\theta-L_1 \cos\alpha -D_{\rm SN}\,,
    \label{eq:deltat}
\end{align}
with $D_{\rm SN}$ the distance between the SN and the Earth, and $L_1$ the distance mother particles propagated before decaying.

The flux of daughter $\nu$ is obtained by considering decays that only occur at $R^{\nu}_{\rm SN}\leq L_1 \leq L^{\rm max}_1$. As the smallest decay distance beyond which the daughter neutrino can escape the explosion unperturbed, we take $R^{\nu}_{\rm SN} = \unit[30]{km}$.
$L^{\rm max}_1$ corresponds to the distance $L_1$ with the largest time delay $\delta t$ \cite{Brdar:2023tmi}. For $\delta t = \unit[100]{s}$ while larger values of $\delta t$ do not affect the limits obtained, we show the differential fluxes of daughter neutrino in comparison to the standard neutrino flux in \cref{fig:fluxes} at different times for two benchmarks 
\begin{figure}[t!]
    \centering
    \includegraphics[width=0.47\textwidth]{figures/majoran_fluxes.pdf}
    \caption{\textbf{\textit{Neutrino flux from Standard Model production and two majoran hypotheses.}}
    Jeff needs to write here.
    }
    \label{fig:fluxes}
\end{figure}

>>>>

\textbf{\textit{The Models}}---
As an example, we will consider the Majoron model. Nevertheless the study can also be applied to sterile neutrinos via magnetic portal, as we discuss in the Supplementary Materials. For the case of sterile neutrino with mixing portal, special care needs to be taken for resonant conversions in which case active neutrinos with plasma mass would travel as slowly as the sterile neutrinos at specific distances.
The relevant part of the Lagrangian in the Majoron case reads
\begin{align}
\mathcal{L} \supset -\frac{g_{\alpha\beta}}{2} \nu_\alpha \nu_\beta \phi - m_\phi \phi\phi^* + \text{h.c.}\,,
\end{align}
where $m_\phi$ is the Majoron mass and $g$ parametrizes interaction strength between $\phi$ and neutrinos. Consider flavor universal interaction, we abbreviate $g_{\alpha\beta}\equiv g$. Majorons are produced from neutrino coalescence in the star and then subsequently decay to a pair of neutrinos with the total decay width to all flavor of neutrinos given by $\Gamma_\phi = 3g^2 m_\phi/16 \pi$, and as a result, BSM neutrino flux is produced.

Using data simulated by the Garching group for an $8.8 M_\odot$ progenitor star \cite{Huedepohl2010}, we obtained the fluxes of standard neutrino case that can be fitted to the parametrization in \cite{baxter2021snewpy} and the fluxes of Majoron $\phi$. Requiring the energy loss to Majorons to be less than 10\% of the standard neutrino energy, we can produce the cooling limit, which is in agreement with those obtained in literature \cite{Fiorillo:2022cdq}.
As mentioned above, for the regions below the coupling strength, Majorons produced will immediately stream out. Along its way towards the Earth, Majorons will decay to active neutrinos. For a given emission angle $\alpha$ of daughter neutrino relative to the direction of mother particle, and the angle $\theta$ at which the daughter neutrino arrives at the Earth, the time delay $\Delta t$, the arrival time of the daughter $\nu$ relative to that of the neutrinos produced in the explosion via SM processes, is \cite{Jaeckel:2017tud}
\begin{align}
    \Delta t = L_1/\beta + D_{\rm SN} \cos\theta-L_1 \cos\alpha -D_{\rm SN}\,,
    \label{eq:deltat}
\end{align}
with $D_{\rm SN}$ the distance between the SN and the Earth, $L_1$ the distance mother particles propagated before decaying, and $\beta$ being the speed of $\phi$. The flux of daughter $\nu$ is obtained by considering decays that only occur at $R^{\nu}_{\rm SN}\leq L_1 \leq L^{\rm max}_1$. As the smallest decay distance beyond which the daughter neutrino can escape the explosion unperturbed, we take $R^{\nu}_{\rm SN} = \unit[30]{km}$.
$L^{\rm max}_1$ corresponds to the distance $L_1$ with the largest time delay $\delta t$ \cite{Brdar:2023tmi}.

\begin{figure}[t!]
    \centering
    \includegraphics[width=0.47\textwidth]{figures/majoran_fluxes.pdf}
    \caption{\textit{\textbf{Neutrino flux from Standard Model production and two majoron hypotheses}.
    We choose the parameters $g_\phi = 10^{-10.2}$ and $g_\phi = 10^{-11.7}$ for $m_\phi = \unit[1]{MeV}$ and $m_\phi = \unit[100]{MeV}$, respectively.}
    }
    \label{fig:fluxes}
\end{figure}

For $D_{\rm SN} = \unit[10]{kpc}$ and $\delta t = \unit[100]{s}$ while larger values of $\delta t$ do not affect the limits obtained, we show in \cref{fig:fluxes} the differential fluxes of daughter neutrinos at different time for two benchmarks of the Majoron case, in comparison to the standard neutrino flux.
We can observe that the energy of neutrinos from Majoron decay extends to $\unit[100]{MeV}$ and above, as expected. For heavy Majoron case, the resulting neutrino flux typically is larger at later time, manifesting the time delay from slowly moving Majorons. 
The flux for light Majoron case contain very rich physics. Majorons, generating neutrinos with $E_\nu \geq \unit[150]{MeV}$, are nearly-relativistic with its daughter particles moving in parallel. In such a case, for couplings comparable to or larger than the one shown in \cref{fig:fluxes}, $\Delta t \sim L_d (1/\beta -1)\lesssim \unit[0.01]{sec}$, with $L_d = E_\phi/m_\phi \Gamma^{-1}_\phi \beta$ being the decay length of $\phi$ for the given $\phi$ energy $E_\phi$. The time dependence of neutrino flux thus are mainly inherited from the Majorons upon production, which exhibits in \cref{fig:fluxes} larger fluxes from producing Majorons from the denser and hotter core of SN at earlier time.
For neutrinos with lower energy, i.e $E_\nu \lesssim \unit[40]{MeV}$, the mother Majorons can be non-relativistic, resulting in $\Delta t \gtrsim \unit[0.04]{sec}$. This non-negligible time delay would shift the neutrino flux to later time, lowering the fluxes at earlier time as shown in \cref{fig:fluxes}. For the heavy mass benchmark in \cref{fig:fluxes}, we have $\Delta t \gtrsim 10 sec$, which would yield the active neutrino flux at sizeable $t\sim \mathcal{sec}$.

%in details in the following, we consider energy deposits at the photomultipliers, 
%due to denser and hotter core of SN  as Majorans shortly after the bouncing time $t_{\rm bounce}$ from the hot dense core of SN.

\begin{figure*}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/hits_and_likelihood.pdf}
    \caption{\textbf{\textit{Test statistic as a function of $t_{\mathrm{start}}$ and $\Delta$t.}
    }}
    \label{fig:hits_and_likelihood}
\end{figure*}

### 2024-01-18 paper/sections/preamble.tex
Updates from Overleaf

\preprint{APS/123-QED, USTC-ICTS/PCFT-24-xx}

\title{Time Prints of New Physics at Neutrino Telescopes
%Supernova Detection with Neutrino Telescopes as a Portal to New Physics
}

\author{Jeff Lazar}
\email{jlazar@icecube.wisc.edu}
\affiliation{Department of Physics \& Laboratory for Particle Physics and Cosmology, Harvard University, Cambridge, MA 02138, USA}
\affiliation{Department of Physics \& Wisconsin IceCube Particle Astrophysics Center, University of Wisconsin-Madison, Madison, WI 53706, USA}
\author{Ying-Ying Li}
\email{yingyingli@ustc.edu.cn}
\affiliation{Peng Huanwu Center for Fundamental Theory, Hefei, Anhui 230026, China}
\affiliation{Interdisciplinary Center for Theoretical Study, University of Science and Technology of China, Hefei, Anhui 230026, China}
\author{Carlos A. Arg\"{u}elles}
\email{carguelles@g.harvard.edu}
\affiliation{Department of Physics \& Laboratory for Particle Physics and Cosmology, Harvard University, Cambridge, MA 02138, USA}
\author{Vedran Brdar}
\email{vedran.brdar@cern.ch}
\affiliation{Department of Physics, Oklahoma State University, Stillwater, OK, 74078, USA}
\affiliation{Theoretical Physics Department, CERN, Esplande des Particules, 1211 Geneva 23, Switzerland}

\date{\today}

\begin{abstract}
More than 30 years ago, a handful of neutrino events from SN 1987A was recorded in several 
neutrino experiments. In this decade, several large-scale neutrino experiments will operate and be able to detect thousands of neutrino events from the next galactic supernova (SN). Such detection would not only lead to a great progress in our understanding of the astrophysical properties of SNe, but would also have a great importance in testing light dark sector models. Traditionally, the new physics constraints from SNe have been associated with SN energy loss arguments. Recently, it was shown that if light new physics states streaming out of the SN decay to neutrinos, non-observation of additional events at present and near future neutrino experiments would lead to limits that surpass the energy loss ones. In this work, focusing on the two representative new physics scenarios, we perform sensitivity projections for new physics searches at several neutrino telescopes, in particular IceCube, where a galactic SN event would be observed through the sharp rise in  photomultiplier counts. Based on these counts, new physics induced events can be constrained. The derived sensitivities strongly exceed constraints from both energy loss and non-observation of additional neutrino events from SN 1987A. We also release a code that allows for efficient computation of sensitivities for new physics searches at neutrino telescopes in the context of galactic SN event. An arbitrary light dark sector model can be studied and the only requirement is that the decay of a light dark sector particle(s) to neutrinos is realized.      
\end{abstract}

>>>>

\preprint{APS/123-QED, USTC-ICTS/PCFT-24-xx}
\bibliographystyle{apsrev4-1}
\title{
% New Physics in Supernovae at Neutrino Telescopes
% Time Prints of New Physics from Supernovae at Neutrino Telescopes
% Supernova Detection with Neutrino Telescopes as a Portal to New Physics
Supernovae Time Profiles as a Portal to New Physics at Neutrino Telescopes
}

\author{Jeff Lazar}
\email{jlazar@icecube.wisc.edu}
\affiliation{Department of Physics \& Laboratory for Particle Physics and Cosmology, Harvard University, Cambridge, MA 02138, USA}
\affiliation{Department of Physics \& Wisconsin IceCube Particle Astrophysics Center, University of Wisconsin-Madison, Madison, WI 53706, USA}
\author{Ying-Ying Li}
\email{yingyingli@ustc.edu.cn}
\affiliation{Peng Huanwu Center for Fundamental Theory, Hefei, Anhui 230026, China}
\affiliation{Interdisciplinary Center for Theoretical Study, University of Science and Technology of China, Hefei, Anhui 230026, China}
\author{Carlos A. Arg\"{u}elles}
\email{carguelles@g.harvard.edu}
\affiliation{Department of Physics \& Laboratory for Particle Physics and Cosmology, Harvard University, Cambridge, MA 02138, USA}
\author{Vedran Brdar}
\email{vedran.brdar@okstate.edu}
\affiliation{Department of Physics, Oklahoma State University, Stillwater, OK, 74078, USA}

%\date{\today}

\begin{abstract}
Neutrino telescopes, such as IceCube, have promised precise time resolutions at the millisecond level for detecting collective rise signals in all photomultiplier.
We demonstrate that such precise time resolutions will play a crucial role in tracking neutrino signals with novel timing patterns related to weakly interacting massive particles in physics beyond the Standard Model.
Focusing on future Supernova explosion in the operation time of IceCube, Focusing on a future Supernova explosion in the operation time of IceCube as an example, we show that neutrino telescopes can provide unprecedented probes of new particles with both low and high masses, where neutrino signals are advanced and delayed, respectively, while comparable to DUNE and Hyper-K in the intermediate mass region.
\end{abstract}
